---
title: "model1 custom GSVA KC-adj fit DEGs"
author: "Lewis E Tomalin, PhD"
date: "2025-06-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      eval = FALSE,
                      message = FALSE,
                      warning = FALSE)

require(tableone)
require(tidyverse)
require(variancePartition)

main_file <- 'genes_associated_with_canonical_kc_ssgsea/results/vobj_fit_contrasts.rda'

```

## This script identifies genes associated custom ssgsea KC scores

Only MIS, RGP and VGP were used

FCH > 1, FDR < 0.05

+ numbers of DEGs associated with custom GSVA KC 
+ volcano plots
+ heatmaps
+ Hallmark and Reactome enrichment

```{r}

require(tableone)
require(tidyverse)
library(variancePartition)

load('data/processed/DGE_final_samples.rda')

dge <- dge_final

load(file='data/processed/db_ssgsea_custom_correct.rda')


dge$samples

dge$samples$ssgsea_canonical_kc <- db_ssgsea_custom_correct[rownames(dge$samples), 'ssgsea_canonical_kc']


dge <- calcNormFactors(dge)

design <- model.matrix(~ ssgsea_canonical_kc + sex + age_centered_impute, 
                       data = dge$samples)

keep <- filterByExpr(dge, 
                     design = design)

table(keep)

dge <- dge[keep, , keep.lib.sizes = FALSE]

dge <- calcNormFactors(dge)


```


```{r echo=T}

# Define formula
form <- ~ ssgsea_canonical_kc + sex + age_centered_impute + (1 | patient_id/final_tissue_type_category)

```


```{r}

# Define formula
form <- ~ ssgsea_canonical_kc + sex + age_centered_impute + (1 | patient_id/final_tissue_type_category)


# Optional: speed up computation
#register(SnowParam(20)) # or use SnowParam for Windows
#bpparam()

# For Mac
library(BiocParallel)
#bp <- MulticoreParam(8) 
bp <- SnowParam(20) 

## Need to explore missing a little more...

dge$samples %>% dim()

vobj <- voomWithDreamWeights(dge, 
                             formula=form, 
                             data = dge$samples,
                             plot = TRUE,
                             save.plot = TRUE,
                             BPPARAM = bp)



fit_contrast <- dream(vobj$E, 
                      formula=form, 
                      data=vobj$targets,
                      BPPARAM = bp)

fit <- variancePartition::eBayes(fit_contrast, trend = F)

save(fit, vobj, dge,
     file=main_file)

register(SerialParam())

```

Analysis of results 

```{r}

load(main_file)

dge$genes -> Annot

head(Annot)

fit$coefficients %>% head()

COEFS <- c('ssgsea_canonical_kc')

L <- 'ssgsea_canonical_kc'

lapply(COEFS, function(L){
  
  # Compare Mel vs Nevus Low
  topTable(fit, 
           coef = L, 
           adjust.method = "fdr",
           p.value = 1,
           number = Inf) %>% 
    mutate(DIRECTION=sign(logFC),
           SIG=case_when(sign(logFC)==1 & adj.P.Val <= 0.05 ~ 'Up',
                         sign(logFC)==-1 & adj.P.Val <= 0.05 ~ 'Down',
                         TRUE ~ 'NS')) %>%
    rownames_to_column('GeneID') %>% 
    mutate(contrast=L) -> temp
  
}) -> contr_list

contr_list %>%
  reduce(full_join) %>%
  left_join(Annot) %>% 
  #mutate(contrast = case_when(contrast == 'csd_continous:tissue_type_simpleMIS' ~ 'MISvsNevus',
  #                            contrast == 'csd_continous:tissue_type_simpleMel' ~ 'MelvsNevus',
  #                            TRUE ~ contrast)) %>%
  mutate(contrast=factor(contrast)) -> db_contrasts

db_contrasts %>%
  group_by(SIG, contrast) %>%
  summarise(COUNT=n()) %>% 
  pivot_wider(id_cols = contrast,
              names_from = SIG,
              values_from = COUNT) -> N_DEGs

N_DEGs %>% as.data.frame()

db_contrasts %>%
  filter(SIG == 'Up') %>%
  arrange(desc(logFC)) %>%
  group_by(contrast) %>%
  slice_head(n = 20) %>%
  mutate(Top10=Symbol) %>%
  select(SIG, contrast, Symbol, Top10) -> Top10_DEGs_Up

db_contrasts %>%
  filter(SIG == 'Down') %>%
  arrange(logFC) %>%
  group_by(contrast) %>%
  slice_head(n = 20) %>% 
  mutate(Top10=Symbol) %>%
  select(SIG, contrast, Symbol, Top10) -> Top10_DEGs_Down

db_contrasts %>% 
  full_join(bind_rows(Top10_DEGs_Up, Top10_DEGs_Down)) -> db_contrasts

save(fit, vobj, db_contrasts, N_DEGs, dge,
     file=main_file)

```


```{r}

library(fgsea)
library(msigdbr)
library(tidyverse)

load(main_file)

hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H")

colnames(hallmark_sets)

head(hallmark_sets)  

hallmark_sets$gs_name %>% unique()

hallmark_sets %>% str()

hallmark_sets %>% 
  filter(gs_name == 'HALLMARK_MYOGENESIS')

# Create a list where each element is a vector of genes for a pathway
hallmark_list <- split(hallmark_sets$gene_symbol, hallmark_sets$gs_name)

L = 'Keratinocytes'

lapply(unique(db_contrasts$contrast), function(L){

db_contrasts %>% 
  filter(contrast == L,
         complete.cases(Symbol)) -> dream_results

# Example: using t-statistic
ranks <- dream_results %>%
  dplyr::filter(!is.na(t)) %>%
  dplyr::arrange(desc(t)) %>%
  dplyr::select(Symbol, t)

# Create a named vector for fgsea
ranked_genes <- setNames(ranks$t, ranks$Symbol)

fgsea_res <- fgsea(pathways = hallmark_list,
                   stats = ranked_genes,
                   minSize = 15,
                   maxSize = 500,
                   nproc = 1)

# Top pathways
fgsea_res <- fgsea_res[order(fgsea_res$pval), ]

fgsea_res$contrast <- L

return(fgsea_res)

}) %>% 
  bind_rows() -> fgsea_res_HK

save(fit, vobj, db_contrasts, N_DEGs, fgsea_res_HK,
    file=main_file)

# load('results/model_6_custom GSVAKC_adjust/vobj_fit_contrasts.rda')
# 
write.csv(fgsea_res_HK %>% 
             select(-leadingEdge), 
           file="genes_associated_with_canonical_kc_ssgsea/results/fgsea_hallmark_results.csv", 
          row.names = FALSE)

```



```{r}

load(file=main_file)

reactome_sets <- msigdbr(species = "Homo sapiens", 
                         collection = "C2") %>%
  filter(str_detect(gs_name, 'REACTOME')) %>%
  dplyr::select(gs_name, gene_symbol)

# Create a list where each element is a vector of genes for a pathway
reactome_list <- split(reactome_sets$gene_symbol, reactome_sets$gs_name)

L <- "Mel_vs_Nevus"

lapply(unique(db_contrasts$contrast), function(L){
  
  db_contrasts %>% 
    filter(contrast == L,
           complete.cases(Symbol)) -> dream_results
  
  # Example: using t-statistic
  ranks <- dream_results %>%
    dplyr::filter(!is.na(t)) %>%
    dplyr::arrange(desc(t)) %>%
    dplyr::select(Symbol, t)
  
  # Create a named vector for fgsea
  ranked_genes <- setNames(ranks$t, ranks$Symbol)
  
  fgsea_res <- fgsea(pathways = reactome_list,
                     stats = ranked_genes,
                     minSize = 15,
                     maxSize = 500,
                     nproc = 8)
  
  # Top pathways
  fgsea_res <- fgsea_res[order(fgsea_res$pval), ]
  
  fgsea_res$contrast <- L
  
  return(fgsea_res)
  
}) %>% 
  bind_rows() -> fgsea_res_REACTOME


save(fit, vobj, db_contrasts, N_DEGs, fgsea_res_HK, fgsea_res_REACTOME,
     file=main_file)

#load('results/model_6_custom GSVAKC_adjust/vobj_fit_contrasts.rda')

write.csv(fgsea_res_REACTOME %>% 
            select(-leadingEdge), 
          file="genes_associated_with_canonical_kc_ssgsea/results/fgsea_reactome_results.csv", 
          row.names = FALSE)



```

```{r eval=T, fig.width=5.5, fig.height=4.5}

load(main_file)

plot(vobj$voom.xy, xlab = "Log2 CPM", ylab = "√(standard deviation)", 
     main = "Mean-variance trend",
     pch = 16, cex = 0.2, col = "black")
lines(vobj$voom.line, col = "red", lwd = 2)

```


## Volcano plots, genes associated with custom GSVA KC KC proportion

```{r echo=T}

form <- ~ ssgsea_canonical_kc + sex + age_centered_impute + (1 | patient_id/final_tissue_type_category)

```

```{r eval=T, fig.width=9, fig.height=6}

load(main_file)

N_DEGs %>% as.data.frame()

db_contrasts %>%
  filter(contrast != 'DeltaVGP_MIS_VS_DeltaVGP_RGP') %>%
  ggplot(aes(x=logFC,
             y=-log10(adj.P.Val),
             color=SIG,
             alpha=SIG))+
  geom_point(size=0.2)+
  ggrepel::geom_text_repel(aes(label=Top10), max.overlaps=20, show.legend = F, color='black')+
  scale_alpha_manual(values=c(Up=1, Down=1, NS=0.1))+
  scale_color_manual(values=c(Up='red', Down='blue', NS='gray'))+
  geom_hline(yintercept=-log10(0.05), linetype='dotted')+
  geom_vline(xintercept=c(-1, 1), linetype='dotted')+
  facet_wrap(~contrast, nrow=3, scales='free_y')+
  labs(title=paste0('Gene associated with custom GSVA Keratinocyte score, FDR < 0.05, Top20'),
       x='strength of accociation with custom GSVA KC score')+
  theme_bw() +
  theme(legend.position = 'none') -> p1

p1


```


## Heatmaps

```{r eval=T, fig.width=12, fig.height=80}

load(main_file)

vobj$genes -> Annot

AnnotVec <- Annot$Symbol
names(AnnotVec) <- Annot$GeneID

require(pheatmap)

db_contrasts %>%
  filter(SIG !='NS',
         complete.cases(Symbol)) %>% 
  .$GeneID %>% unique() -> AllDEGs

vobj$targets -> PHENO

PHENO %>% 
  arrange(final_tissue_type_category, ssgsea_canonical_kc, patient_id) -> PHENO

require(RColorBrewer)

BREAKS <- seq(-2, 2, by=0.1)

pheatmap(vobj$E[AllDEGs, rownames(PHENO)],
         breaks = BREAKS,
         color = colorRampPalette(rev(brewer.pal(n = 8, name ="RdBu")))(length(BREAKS)),
         show_rownames = F, 
         show_colnames = F,
         #labels_row = AnnotVec[AllDEGs],
         annotation_col=PHENO %>% select(final_tissue_type_category, ssgsea_canonical_kc),
         annotation_colors = list(final_tissue_type_category=c(Nevus='green',
                                                               Intermediate='yellow',
                                                               MIS='orange',
                                                               RGP='red',
                                                               VGP='purple')),
         cluster_cols = F,
         cluster_rows = T,
         #cellwidth = 5,
         #cellheight = 10,
         main='All genes associated with custom ssGSEA KC KC proportion',
         gaps_col = which(!duplicated(PHENO$final_tissue_type_category))[-1]-1,
         scale = 'row')



```

## Heatmap by KC proportion

```{r eval=T, fig.width=12, fig.height=80}

load(main_file)

vobj$genes -> Annot

AnnotVec <- Annot$Symbol
names(AnnotVec) <- Annot$GeneID

require(pheatmap)

db_contrasts %>%
  filter(SIG !='NS',
         complete.cases(Symbol)) %>% 
  .$GeneID %>% unique() -> AllDEGs

vobj$targets -> PHENO

PHENO %>% 
  arrange(ssgsea_canonical_kc, patient_id) -> PHENO

require(RColorBrewer)

BREAKS <- seq(-2, 2, by=0.1)

pheatmap(vobj$E[AllDEGs, rownames(PHENO)],
         breaks = BREAKS,
         color = colorRampPalette(rev(brewer.pal(n = 8, name ="RdBu")))(length(BREAKS)),
         show_rownames = F, 
         show_colnames = F,
         #labels_row = AnnotVec[AllDEGs],
         annotation_col=PHENO %>% select(final_tissue_type_category, ssgsea_canonical_kc),
         annotation_colors = list(final_tissue_type_category=c(Nevus='green',
                                                               Intermediate='yellow',
                                                               MIS='orange',
                                                               RGP='red',
                                                               VGP='purple')),
         cluster_cols = F,
         cluster_rows = T,
         #cellwidth = 5,
         #cellheight = 10,
         main='All genes associated with custom ssGSEA KC score',
         #gaps_col = which(!duplicated(PHENO$final_tissue_type_category))[-1]-1,
         scale = 'row')



```


## GSEA analysis



```{r eval=T, fig.width=10, fig.height=12}

load(file=main_file)

fgsea_res_HK %>%
  filter(padj <= 0.1) %>% 
  .$pathway %>% 
  unique() -> SIG_PATH

fgsea_res_HK %>% 
  filter(pathway %in% SIG_PATH) %>%
  select(-leadingEdge) %>%
  pivot_wider(id_cols = pathway,
              values_from = NES,
              names_from = contrast) %>%
  mutate(pathway=str_remove(pathway, 'HALLMARK_')) %>%
  column_to_rownames('pathway') -> mat

fgsea_res_HK %>% 
  rstatix::add_significance(p.col = 'padj',
                            cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                            symbols = c("****", "***", "**", "*", "")) %>%
  filter(pathway %in% SIG_PATH) %>%
  select(-leadingEdge) %>%
  pivot_wider(id_cols = pathway,
              values_from = padj.signif,
              names_from = contrast) %>%
  mutate(pathway=str_remove(pathway, 'HALLMARK_')) %>%
  column_to_rownames('pathway') -> STARS

BREAKS <- seq(-3, 3, by=0.1)

require(RColorBrewer)
require(pheatmap)

#dev.off()
#pdf('results/model_1_melanoma/gsea_stuff/HallmarkHeatmap.pdf', width=8, height=12)
pheatmap(mat,
         breaks = BREAKS,
         color = colorRampPalette(rev(brewer.pal(n = 8, name ="RdBu")))(length(BREAKS)),
         show_rownames = T, 
         show_colnames = T,
         display_numbers = STARS,
         cluster_cols = F,
         cellheight = 12,
         cellwidth = 25,
         main='Hallmark Enrichment, FDR < 0.1')
#dev.off()

```



### Reactome Top 5 up and down (per contrast)


```{r eval=T, fig.width=10, fig.height=12}


load(file=main_file)

## Reactome Heatmap

fgsea_res_REACTOME %>%
  filter(padj <= 0.1) %>% 
  arrange(NES) %>%
  group_by(contrast) %>%
  slice_head(n=10) %>%
  .$pathway %>% 
  unique() -> SIG_PATH_DOWN

fgsea_res_REACTOME %>%
  filter(padj <= 0.1) %>% 
  arrange(desc(NES)) %>%
  group_by(contrast) %>%
    slice_head(n=10) %>%
  .$pathway %>% 
  unique() -> SIG_PATH_UP

SIG_PATH <- unique(c(SIG_PATH_UP, SIG_PATH_DOWN))

fgsea_res_REACTOME %>% 
  filter(pathway %in% SIG_PATH) %>%
  select(-leadingEdge) %>%
  pivot_wider(id_cols = pathway,
              values_from = NES,
              names_from = contrast) %>%
  mutate(pathway=str_remove(pathway, 'REACTOME_')) %>%
  column_to_rownames('pathway') -> mat

fgsea_res_REACTOME %>% 
  rstatix::add_significance(p.col = 'padj',
                            cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                            symbols = c("****", "***", "**", "*", "")) %>%
  filter(pathway %in% SIG_PATH) %>%
  select(-leadingEdge) %>%
  pivot_wider(id_cols = pathway,
              values_from = padj.signif,
              names_from = contrast) %>%
  mutate(pathway=str_remove(pathway, 'REACTOME_')) %>%
  column_to_rownames('pathway') -> STARS

BREAKS <- seq(-3, 3, by=0.1)

require(RColorBrewer)

#dev.off()
#pdf('results/model_1_melanoma/gsea_stuff/ReactomeHeatmap.pdf', width=16, height=10)
pheatmap(mat,
         breaks = BREAKS,
         color = colorRampPalette(rev(brewer.pal(n = 8, name ="RdBu")))(length(BREAKS)),
         show_rownames = T, 
         show_colnames = T,
         display_numbers = STARS,
         cluster_cols = F,
         cellheight = 12,
         cellwidth = 25,
         main='Reactome Enrichment, Top 10')
#dev.off()

```


### Reactome all


```{r eval=T, fig.width=16, fig.height=40}

load(file=main_file)

## Reactome Heatmap

fgsea_res_REACTOME %>%
  filter(padj <= 0.1) %>% 
  arrange(NES) %>%
  group_by(contrast) %>%
  .$pathway %>% 
  unique() -> SIG_PATH_DOWN

fgsea_res_REACTOME %>%
  filter(padj <= 0.1) %>% 
  arrange(desc(NES)) %>%
  group_by(contrast) %>%
  .$pathway %>% 
  unique() -> SIG_PATH_UP

SIG_PATH <- unique(c(SIG_PATH_UP, SIG_PATH_DOWN))

fgsea_res_REACTOME %>% 
  filter(pathway %in% SIG_PATH) %>%
  select(-leadingEdge) %>%
  pivot_wider(id_cols = pathway,
              values_from = NES,
              names_from = contrast) %>%
  mutate(pathway=str_remove(pathway, 'REACTOME_')) %>%
  column_to_rownames('pathway') -> mat

fgsea_res_REACTOME %>% 
  rstatix::add_significance(p.col = 'padj',
                            cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                            symbols = c("****", "***", "**", "*", "")) %>%
  filter(pathway %in% SIG_PATH) %>%
  select(-leadingEdge) %>%
  pivot_wider(id_cols = pathway,
              values_from = padj.signif,
              names_from = contrast) %>%
  mutate(pathway=str_remove(pathway, 'REACTOME_')) %>%
  column_to_rownames('pathway') -> STARS

BREAKS <- seq(-3, 3, by=0.1)

require(RColorBrewer)

#dev.off()
#pdf('results/model_1_melanoma/gsea_stuff/ReactomeHeatmap.pdf', width=16, height=10)
pheatmap(mat,
         breaks = BREAKS,
         color = colorRampPalette(rev(brewer.pal(n = 8, name ="RdBu")))(length(BREAKS)),
         show_rownames = T, 
         show_colnames = T,
         display_numbers = STARS,
         cluster_cols = F,
         cellheight = 12,
         cellwidth = 25,
         main='Reactome Enrichment')
#dev.off()

```


```{r eval=F, fig.width=13, fig.height=8}


load(file=main_file)

# Plot a specific enrichment
#plotEnrichment(reactome_list[["REACTOME_NONSENSE_MEDIATED_DECAY_NMD"]], ranked_genes)

#fgsea_res_all %>% 
#  select(-leadingEdge) %>%
#  filter(pathway=="REACTOME_NONSENSE_MEDIATED_DECAY_NMD") %>%
#  arrange(NES)

fgsea_res_REACTOME %>% 
  filter(pathway=="REACTOME_NONSENSE_MEDIATED_DECAY_NMD") %>% .$leadingEdge -> Ledge

Ledge %>%
  reduce(unique) -> Ledge

require(msigdbr)

reactome_sets <- msigdbr(species = "Homo sapiens", 
                         collection = "C2") %>%
  filter(str_detect(gs_name, 'REACTOME')) %>%
  dplyr::select(gs_name, gene_symbol)

# Create a list where each element is a vector of genes for a pathway
reactome_list <- split(reactome_sets$gene_symbol, reactome_sets$gs_name)

reactome_list$REACTOME_NONSENSE_MEDIATED_DECAY_NMD -> PTWY

## Volcano plots

db_contrasts %>%
  filter(contrast != 'DeltaVGP_MIS_VS_DeltaVGP_RGP') %>%
  #mutate(adj.P.Val=case_when(contrast == 'RGP_vs_MIS' & adj.P.Val < 0.5 ~ 1,
  #                           TRUE ~ adj.P.Val)) %>%
  mutate(LeadingEdge=case_when(Symbol %in% Ledge ~ 'LeadingEdge',
                               Symbol %in% PTWY ~ 'Pathway',
                               TRUE ~ 'No'),
         Label=case_when(LeadingEdge == 'LeadingEdge' ~ Symbol,
                               TRUE ~ NA_character_)) %>%
  mutate(LeadingEdge = factor(LeadingEdge, levels = c("No", "Pathway", "LeadingEdge"))) %>%
  arrange(LeadingEdge) %>% 
  ggplot(aes(x=logFC,
             y=-log10(adj.P.Val),
             color=LeadingEdge))+
  geom_point(aes(size=LeadingEdge))+
  #ggrepel::geom_text_repel(aes(label=Label), max.overlaps=20, show.legend = F, color='black')+
  #scale_alpha_manual(values=c(Up=1, Down=1, NS=0.1))+
  scale_size_manual(values=c(LeadingEdge=1, Pathway=1, No=0.2))+
  scale_color_manual(values=c(LeadingEdge='forestgreen', Pathway='blue', No='gray'))+
  #geom_hline(yintercept=-log10(0.05), linetype='dotted')+
  geom_vline(xintercept=c(-1, 1), linetype='dotted')+
  facet_wrap(~contrast, scales='free')+
  labs(title=paste0('NMD genes highlighted'))+
  theme_bw() 


```


### Volcano plots with KRT genes highlighted

```{r eval=T, fig.width=6, fig.height=4}


load(file=main_file)

# Plot a specific enrichment
#plotEnrichment(reactome_list[["REACTOME_NONSENSE_MEDIATED_DECAY_NMD"]], ranked_genes)


## Volcano plots

db_contrasts %>%
  filter(contrast != 'DeltaVGP_MIS_VS_DeltaVGP_RGP') %>%
  #mutate(adj.P.Val=case_when(contrast == 'RGP_vs_MIS' & adj.P.Val < 0.5 ~ 1,
  #                           TRUE ~ adj.P.Val)) %>%
  mutate(KRT=case_when(str_detect(Symbol, 'KRT')~ 'KRT gene',
                               TRUE ~ 'not KRT'),
         Label=case_when(KRT == 'KRT gene' ~ Symbol,
                               TRUE ~ NA_character_)) %>%
  arrange(desc(KRT)) %>% 
  ggplot(aes(x=logFC,
             y=-log10(adj.P.Val),
             color=KRT))+
  geom_point(aes(size=KRT))+
  #ggrepel::geom_text_repel(aes(label=Label), max.overlaps=20, show.legend = F, color='black')+
  #scale_alpha_manual(values=c(Up=1, Down=1, NS=0.1))+
  scale_size_manual(values=c(`KRT gene`=1, `not KRT`=0.2))+
  scale_color_manual(values=c(`KRT gene`='forestgreen', `not KRT`='gray'))+
  geom_hline(yintercept=-log10(0.05), linetype='dotted')+
  geom_vline(xintercept=c(-1, 1), linetype='dotted')+
  facet_wrap(~contrast, scales='free', nrow=1)+
  labs(title=paste0('KRT genes highlighted'))+
  theme_bw() 


```


### Volcano plots from tissue_type_degs with xcell KC-associated genes highlighted

```{r eval=T, fig.width=13, fig.height=12}


load(file='genes_associated_with_canonical_kc_ssgsea/results/vobj_fit_contrasts.rda')

db_contrasts %>%
  filter(SIG == 'Up') %>%
  arrange(desc(logFC)) %>%
  select(SIG, contrast, Symbol, Top10) %>%
  .$Symbol -> canonical_kc_associated_genes

load(file='tissue_type_degs/results/vobj_fit_contrasts.rda')

## Volcano plots

db_contrasts %>%
  filter(contrast != 'DeltaVGP_MIS_VS_DeltaVGP_RGP') %>%
  mutate(canon_kc_associated=case_when(Symbol %in% canonical_kc_associated_genes ~ 'canon kc assoc',
                                       TRUE ~ 'not assoc'),
         Label=case_when(str_detect(Symbol, 'KRT') ~ Symbol,
                         TRUE ~ NA_character_)) %>%
  arrange(desc(canon_kc_associated)) %>% 
  ggplot(aes(x=logFC,
             y=-log10(adj.P.Val),
             color=canon_kc_associated))+
  geom_point(aes(size=canon_kc_associated))+
  ggrepel::geom_text_repel(aes(label=Label), max.overlaps=20, show.legend = F, color='black')+
  #scale_alpha_manual(values=c(Up=1, Down=1, NS=0.1))+
  scale_size_manual(values=c(`canon kc assoc`=1, `not assoc`=0.2))+
  scale_color_manual(values=c(`canon kc assoc`='forestgreen', `not assoc`='gray'))+
  #geom_hline(yintercept=-log10(0.05), linetype='dotted')+
  geom_vline(xintercept=c(-1, 1), linetype='dotted')+
  facet_wrap(~contrast, scales='free', nrow=4)+
  labs(title=paste0('canon kc genes highlighted'))+
  theme_bw() 


```

